Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-14 transmission-appendix-phase-type.tex
add INLA appendix

\begin{document}
\ifSubfilesClassLoaded{
    \setcounter{chapter}{7}
}

\chapter{Estimating transmission} \label{transmission}

\section{Backcalculation} \label{backcalc}

Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
In this chapter, I will apply the theory introduced in \cref{E-inc-prev} to the CIS data.
I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.

\todo[inline]{
    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
    At the same place, I could review the literature on mechanistic models using REACT and CIS.
}

\subsection{Methods} \label{backcalc:sec:methods}

As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
This is a GLM with a logit link function.
The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.

I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
The model included fixed effects for age group\todo{check the model I used}.
The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.

I used age groupings:\todo{insert age groupings}.
These align with previous work on the CIS\todo{cite CIS reports}.
They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
It is lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

\subsection{Results} \label{backcalc:sec:results}

\subsection{Discussion} \label{backcalc:sec:discussion}

\section{Mechanistic modelling}

{\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
\begin{itemize}
    \item 8.3 (currently  titled Implementation) CIS-based transmission model (see comments on lines 484-486) 
\item 8.3.1 WAIFW matrix
\item 8.3.2 Full model (perhaps another title..model structure?) 
\item 8.3.3 Observational model (currently section 8.2) include here the "Solving the system of ODEs" (currently in section 8.3.4)
\item 8.3.4 Priors 
\item 8.3.5 Inference 
\item 8.4 Simulation study
\end{itemize}
}

\todo[inline]{Redo this intro, with some high-level comments taken from the full model section (eg following Birrell et al.)}

\subsection{Methods}

I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
The model will be stratified only by region and age group, so the data are aggregated to these levels.
Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.

\subsubsection{WAIFW matrix}

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
\todo{Contrast with age groupings used in previous chapter}
\begin{enumerate}
    \item $[0, 11)$: children of pre-school and primary school age.
    \item $[11, 16)$: children of compulsory secondary-school age.
    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
    \item $[25, 50)$: younger working age adults.
    \item $[50, 70)$: older working age adults.
    \item $[70, \infty)$: elderly.
\end{enumerate}

I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.

This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method takes pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are then used to down or up-weight the number of contacts in each setting, and then recombined across settings to form $\kappa_{aa'}$
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

Some behavioural changes (\eg mask wearing) impact transmission.
However, these are not captured by the contact survey.
To allow for these changes and the uncertainty produced in formulating $\matr{\kaapa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}).

% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.

Finally,$q_{aa'}$ needs to be specified.
I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.

%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.

Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

\subsubsection{Prevalence and observation models} \label{SEIR:sec:observation}

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}.
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt }.

A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{SEIR:sec:non-exponential}).

@@

% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
An EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
It is shown schematically in \cref{SEIR:fig:EC}.
Here, $K \dist \GamDist(l-2, \kappa_Y)$, $X_1 \dist \Exponential(\kappa_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\kappa_{X2})$.
% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
Hence, the overall distribution is known as a Erlang-Coxian distribution.

@@

    \path (1) edge node {$\kappa_Y$} (2)
          (2) edge node {$\kappa_Y$} (dots)
          (dots) edge node {$\kappa_Y$} (lm2)
          (lm2) edge node {$\kappa_Y$} (lm1)
          (lm1) edge node {$p_x \kappa_{X1}$} (l)
                edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (absorbing)
          (l) edge node {$\kappa_{X2}$} (absorbing);
\end{tikzpicture}
}
\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a series of compartments. The distribution is the time from entering state 1 and entering the absorbing state. The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \kappa_Y)$, also known as an Erlang distribution. The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
\label{SEIR:fig:EC}
\end{figure}

\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
They then prove that EC distributions have the following properties:
\begin{itemize}
    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all phase-type distributions that well-match $G$.
\end{itemize}

A total of five parameters are required to specify a $l$-phase EC distribution.
\begin{itemize}
    \item $l$, the number of phases.
    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
    \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
\end{itemize}

I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
This is the same duration distribution used in \cref{E-backcalc}.
The parameter estimates are in \cref{SEIR:table:ec-params}.
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\kappa_Y$ & $\kappa_{X1}$ & $\kappa_{X2}$ & $p_x$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

\subsubsection{Model equations and solution} \label{SEIR:sec:full-model}

    \node[state] (S) {$s$};
    \node[state, right=of S] (E1) {$e_1$};
    \node[state, right=of E1] (E2) {$e_2$};
    \node[state, right=of E2] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2] (R) {$r$};

    \path (S) edge node {$\lambda$} (E1)
          (E1) edge node {$\sigma$} (E2)
          (E2) edge node {$\sigma$} (I1)
          (I1) edge node {$\gamma$} (I2)
          (I2) edge node {$\gamma$} (R);

    \node[state, below=3cm of E1] (P0) {$p_0$};
    \node[state, right=of P0] (P1) {$p_1$};
    \node[state, right=of P1] (P2) {$p_2$};
    \node[state, right=of P2] (P3) {$p_3$};
    \node[state, right=of P3, draw=none] (P4) {};

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\kappa_0$} (P1)
          (P1) edge node {$\kappa_Y$} (P2)
          (P2) edge node {$p_x \kappa_{X1}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
          (P3) edge node {$\kappa_{X2}$} (P4);

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
}
  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
  \label{SEIR:fig:full-model}
\end{figure}

The full model combines the elements discussed in this chapter up until now.
The transmission model follows \textcite{birrellRealtime}.
It is a SEIR model with two latent and two infectious states.
The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.

To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.

The observation model is novel, relating the latent transmission described by the model to the CIS data.
It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The system of differential equations for each region is below.
\begin{align}
    \label{SEIR:eq:fullODEs}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) (\vec{i_1}(t) + \vec{i_2}(t)) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t).
\end{align}

\subsubsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$

The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.

\subsubsection{Priors} \label{SEIR:sec:priors}

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining priors had weakly informative priors.

\begin{landscape}
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution & Comment \\
    \hline \\
    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalising towards constant \\
    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising towards binomial
\end{tabular}
\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
\label{SEIR:table:priors}
\end{table}

\begin{table}
\centering
\begin{tabular}{l|lllll}
         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
        \hline
        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
    \end{tabular}
\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination. Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
\label{SEIR:table:immunity-prior}
\end{table}
\end{landscape}

\subsubsection{Inference} \label{SEIR:sec:MCMC}

I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
See \cref{SEIR:MCMC-algorithm} for full details.
\begin{algorithm}
 set $\vec{X_0}$ to an initial value of the parameter vector \;
 $\vec{\mu_0} = \vec{X_0}$ \;
 $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
 $\lambda_0 = 1$ \;
 \For{$i = 1, \dots, M$}{
  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
  update the proportion of proposals accepted so far, $\alpha$ \;
  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
    $\lambda_i = \lambda_{i-1}$ \;
    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
   }{
    $\gamma_i = (i - 200)^{-0.6}$ \;
    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
  }
 }
 \caption{Algorithm for adaptive random-walk Metropolis-Hastings. $\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$. $\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$. $\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%). $\gamma_i$ is the learning rate, which determines how much adaptation occurs. $\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}. Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
 \label{SEIR:MCMC-algorithm}
\end{algorithm}

I transformed parameters with restricted ranges for sampling, to improve efficiency.
Parameters which are constrained to the positive reals ($\sigma_\epsilon$, $\rho$, and $q_c$) were sampled on the log scale.
Parameters which are constrained to the unit interval ($i^+$ and $\vec{\pi}$) were sampled on the logit scale.

>>>>

\begin{document}

\chapter{Creating RT-PCR positive states} \label{transmission:sec:phase-type}

@@

% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
It is shown schematically in \cref{SEIR:fig:EC}.
Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-\pi$ or otherwise (\ie with probability $\pi$) distributed $\Exponential(\omega_{X2})$.
% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
These distributions contain five parameters.
Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $\pi$, in the interval $[0, 1]$; and one is a positive integer, $l$.

The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
Hence, the overall distribution is known as a Erlang-Coxian distribution.

@@

    \path (1) edge node {$\omega_Y$} (2)
          (2) edge node {$\omega_Y$} (dots)
          (dots) edge node {$\omega_Y$} (lm2)
          (lm2) edge node {$\omega_Y$} (lm1)
          (lm1) edge node {$\pi \omega_{X1}$} (l)
                edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (absorbing)
          (l) edge node {$\omega_{X2}$} (absorbing);
\end{tikzpicture}
}
\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a progression through a series of compartments. The distribution is the time from entering state 1 and entering the absorbing state.}
\label{SEIR:fig:EC}
\end{figure}

\Textcite{osogamiClosed} define a phase-type distribution $P$ as \emph{well representing} an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
They then prove that EC distributions have the following properties.
\begin{itemize}
    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$.
        Let $\set{W}' \subseteq \set{W}$ be the set of EC distributions that well represent $G$.
        If $\set{W}$ is non-empty (which is true for most real-life distributions), then $\set{W'}$ is also non-empty; the following properties assume that $\set{W}$ is non-empty.
    \item Let $w$ and $w'$ be the minimum number of phases of any member of $\set{W}$ and $\set{W'}$ respectively.
        Then $w' = w$ or $w' = w + 1$.
        That is, the minimum number of phases for an EC distribution to well represent $G$ is at most one more than the minimum number of phases for any phase-type distribution to well represent $G$.
        Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
    \item A $l$-phase EC distribution $P' \in \set{W'}$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}).
        This is beneficial because finding arbitrary members of $\set{W}$ is challenging.
\end{itemize}

A total of five parameters are required to specify a $l$-phase EC distribution.
\begin{itemize}
    \item $l$, the number of phases.
    \item $\omega_Y$, the rate parameter of the Erlang distribution.
    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
    \item $\pi$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
\end{itemize}
